---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(png = list(type = "cairo")),
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GO-a-GO

<!-- badges: start -->
<!-- badges: end -->

GO-a-GO annotates functional terms that are overrepresented in a given set of
gene pairs. The enrichment of Gene Ontology terms is calculated from a
permutation test for overrepresentation of gene pairs that are associated with a
common term.

## Installation

You can install the development version of GO-a-GO from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ajank/GOaGO")
```

## Example

Let us run GO-a-GO on a set of gene pairs associated with human GM12878-specific
chromatin loops. We can inspect a few rows of the dataset:

```{r dataset}
library(GOaGO)

tail(genePairsGM12878Specific)
```

The column names ending with `1` and `2` refer to the first and second loop
anchors, respectively. The essential columns are: `geneID1` and `geneID2` (gene
identifiers from a database of choice), `pairID` (identifier of a gene pair),
and `loopID` (identifier of a chromatin loop).  As some loop anchors overlapped
Transcription Start Sites of multiple genes, the dataset contains all gene
combinations for these loops.

When running GO-a-GO, we should specify that gene identifiers are from the
ENTREZ database, and use the Bioconductor `org.Hs.eg.db` package as the source
of Gene Ontology annotations for human genes, taking all three subontologies
(Biological Process, Molecular Function, and Cellular Component). We will run
10,000 permutations, and use the default *p*-value cutoff of 0.05 with
Benjamini-Hochberg correction:

```{r example, message=FALSE}
library(org.Hs.eg.db)

# set the number of CPU threads to use
library(BiocParallel)
options(MulticoreParam=MulticoreParam(workers=2))

goago <- GOaGO(genePairsGM12878Specific, keyType = "ENTREZID",
               OrgDb = org.Hs.eg.db, ont = "ALL", numPermutations = 10000L)
```

We can inspect the overrepresented GO terms:

```{r result}
head(goago@result)
```

Note that some of the *p*-values have the value of zero, which means that for
all the randomizations fewer gene pairs were associated with a given term than
in the input dataset.

We can plot the overrepresented GO terms as a dotplot. The x-axis shows the
log~2~ fold change of the fraction of gene pairs sharing the given term; `Ratio`
is the fraction in the input gene pairs, while `BgRatio` is the fraction in
permuted ones. Color indicates the adjusted *p*-value:

```{r dotplot}
DotPlot(goago)
```

Note that by default only the terms associated with at least 5 gene pairs are
shown; you can change this by setting `minTermPairs` to any other value.

We can also see the sampling distributions of numbers of gene pairs sharing each
GO term, obtained for the randomized gene pairs. From these distributions,
empirical *p*-values were calculated:

```{r ridgeplot}
RidgePlot(goago)
```
